---
title: "Analyzing Activity and Sleep Patterns Using Fitbit Data to Inform Health Marketing Strategies"
author: "Chris"
date: "2025-07-04"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Problem Statement
The goal of this case study is to understand users' physical activity and sleep behaviors using Fit-bit tracker data. By identifying trends and patterns, the insights can help guide personalized health marketing strategies and feature design to wear.

##  Dataset Description
Using the Fitbit dataset from Kaggle:
https://www.kaggle.com/arashnic/fitbit

Files used:

* dailyActivity_merged.csv
* sleepDay_merged.csv

These contain:

* Physical activity measures (steps, distance, calories, etc.)
* Sleep behavior (total sleep time, time in bed, etc.)

## Setup & Data Loading
### Install and load necessary packacges
```{r Install and load necessary packages}
install.packages("tidyverse")
library(tidyverse)
install.packages("dplyr")
library(dplyr)
```
### Load dataset
```{r load dataset}
daily_activity <- read.csv("dailyActivity_merged.csv")
sleep_day <- read.csv("sleepDay_merged.csv")

```
### Explore Structure
```{r Explore Structure}
glimpse(daily_activity)
glimpse(sleep_day)

```

## Data Cleaning & Summary Stats
### Check participant count
```{r Check participant count}
n_distinct(daily_activity$Id)  
n_distinct(sleep_day$Id)       

```
### Summary statistics
```{r Summary statistics}
summary(select(daily_activity, TotalSteps, TotalDistance, SedentaryMinutes))
summary(select(sleep_day, TotalSleepRecords, TotalMinutesAsleep, TotalTimeInBed))

```
## Insights:

* Almost all users recorded 1 sleep session per day, as expected.
* Users sleep between 58 and 796 minutes/day, with a median of 433 mins (~7.2 hrs).
* Time in bed is slightly longer—median of 463 mins (~7.7 hrs)—implying a ~30-minute gap between being in bed and actually sleeping.

## Exploratory Data Analysis (EDA)
### Relationship between steps and sedentary time
```{r Relationship between steps and sedentary time}
ggplot(daily_activity, aes(x = TotalSteps, y = SedentaryMinutes)) +
  geom_point() +
  labs(title = "Steps vs. Sedentary Minutes")

```

### Sleep vs. Time in Bed
```{r Sleep vs. Time in Bed}
ggplot(sleep_day, aes(x = TotalMinutesAsleep, y = TotalTimeInBed)) +
  geom_point() +
  labs(title = "Minutes Asleep vs. Time in Bed")
```

## Insight

* More steps generally mean less sedentary time, but with exceptions.
* Many users spend more time in bed than they actually sleep — indicating poor sleep efficiency.


## Merging and Deeper Analysis
### Merge datasets by Id
```{r Merge datasets by Id}
combined_data <- merge(sleep_day, daily_activity, by = "Id")
```
### How many participants in merged set?
```{r}
n_distinct(combined_data$Id)
```

### Sleep vs. Activity

```{r Sleep vs. Activity}
ggplot(combined_data, aes(x = TotalMinutesAsleep, y = TotalSteps)) +
  geom_point() +
  labs(title = "Sleep vs. Steps")

```

## Insight:
There's no strong correlation between sleep and daily steps across this small sample, but outliers (very low or very high sleepers) may show interesting behavior.

## Key Findings
* Sedentary Lifestyle: Users are highly sedentary (over 16 hours/day).
* Sleep Efficiency: Users spend significantly more time in bed than they actually sleep.
* Marketing Insight: Positioning the Fitbit as a sleep efficiency tracker could resonate with poor sleepers; promoting gentle activity goals could appeal to sedentary users.

## Recommendations
* Introduce motivational features for users with low daily activity.
* Develop sleep coaching based on time-in-bed vs. sleep duration.
* Use segment-specific messaging: active users vs. sedentary ones.







